---
title: "Variables and Data Description"
author: "Rena Cohen"
date: "11/15/2020"
output:
  html_document: default
  pdf_document: default
---

```{r include = F}
library(gt)
library(tidyverse)
library(readr)
library(readxl)
clean_data_2 <- readRDS("~/Desktop/STAT 139/stat139_project/clean_data_2.rds")

```

Points to hit: 

- Our y variable comes from a NYT survey
  -explain how/where survey was conducted, how we came up with our aggregate variable, possible problems with it
  
- Our x variables were compiled from a variety of sources, based both on what we easily had access to at the count
y level and what prior research has shown about how different factors impact mask mandates
  - This Gallup poll showed different mask usage by gender, political party, education, and age, so those are all included in our analysis (https://news.gallup.com/poll/315590/americans-face-mask-usage-varies-greatly-demographics.aspx).
  - This poll from the Pew Research center showed similar results, as well as variances based on race, so we included those percentages as well (https://www.pewresearch.org/fact-tank/2020/06/23/most-americans-say-they-regularly-wore-a-mask-in-stores-in-the-past-month-fewer-see-others-doing-it/)
    - The choice to include race as a variable also stemmed from the numerous sources that show coronavirus dispproportionately impacting communities of color (https://www.cdc.gov/mmwr/volumes/69/wr/mm6933e1.htm)
  - This recent paper indicated a difference in mask-wearing in stores between rural and urban areas (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7561164/), so we included population density and measurement on the rural-urban continuum as possible predictors
- Finally, we wanted to determine whether coronavirus-related measures, including number of cases/deaths, growth rate of the virus at the time of the survey, and local/statewide mask mandates had impacts on people's behavior. 
  

```{r echo = F}
Variable_Names <- read_excel("Variable Names.xlsx")

Variable_Names %>%
  gt() %>%
  tab_header(title = md("**Variable Names and Descriptions**"))

```



